Split-Apply-Combine with Collapsing Groups

### 2022-12-30 paper/article.Rnw
writing

In all packages mentioned, the calculation for each group uses data available
within the group. However, there are valid use cases where a group aggregate is
determined using attributes from out-of-group entities. One example where this
occurs is in the area of small area estimation (see e.g.
\citet{molina2015sae}. Here, one wishes to estimate an aggregate for a group,
for example a geographical region, or a detailed population subset, where the
number of (sampled) observations so small that the variance of the estimate
would be unacceptably large. Small area estimation is a collection of
bias-variance trade-offs where one `borrows statistical strength' from
out-of-group records to obtain a stable estimate for groups with few
observations. The out-of-group records can be obtained, for example by
combining the original group with a group of records that are deemed similar in
certain respects. A second area where out-of-group records play a role is in
certain hot-deck imputation methods \citep{andridge2010review}.  In
$k$-nearest-neighbours imputation for example, one finds a set of $k$ donor
records that are preferably in the same group, but this condition may be
relaxed if there are not enough records in the group.  In the \pkg{VIM} package
for \proglang{R} \citep{kowarik2016imputation}, this is controlled by a
combination of the Gower distance and setting conditions on minimal number of
donors. In practice, imputation is often performed via a fall-though scenario,
where one first tries to estimate a model within a group, but if the group is
too small for a reliable estimate of model parameters, the group is enlarged by
combining similar groups, similar to the small-area estimation scenario. A
third example comes from the field of statistical disclosure control
\citep{hundepool2012statistical}. In this field, the so-called $p$\% rule
states that the value in a grouped aggregation may not be determined for more
than $p$ percent by a single entity. This occurs for example, when a certain
economic activity in a geographic region is dominated by a single company. In
such cases the value can either not be published (suppression), or one
publishes an aggregate over larger groups.

Common quality issues
include insufficient records, or too many records with values for certain
variables missing. In such cases it is not uncommon to use data from outside
the target group  to estimate the group aggregate. One example is Small Area
Estimation, where the lack of observations in one group is compensated by using
data from similar but other groups. Another area where such approaches occur is
group-wise imputation. For example, one may wish to impute a group estimate
(say, mean, or median) for missing values. If there are too little observations
to provide an accurate estimate of the group mean, the group may be expanded,
trading bias for variance.

\begin{itemize}
\item Add SDC collapsing as application
\end{itemize}

One approach that frequently occurs in practice is to use a so-called
collapsing scheme. In this approach one attempts to estimate the aggregate for
the most fine-grained grouping first.  If this is not possible, two or more
groups are combined to provide input for estimating a value for the
fine-grained group. The collapsing scheme defines which groups may be combined,
and it may consist of multiple levels of collapsing; in principle to the point
where the whole data set is used to estimate an aggregate value for a single
group.

Collapsing schemes can be the result of an optimisation, but they frequently
result from subject matter as well. For example, in the case of hierarchical
classifications, there is a natural ordering of groups that are similar from a
subject matter point of view (whether that means that they can be considered
representative for each other is another, statistical, question).  Examples
include the NACE classification of economic activities or the ISCO
classification of professions. Both have a tree-like structure that can be used
to combine groups with a common ancestor in the hierarchy.

In this paper we formalise the concept of `collapsing scheme', and give an
algorithm for split-apply-combine operations based on it. Next, we demonstrate
the R package \texttt{accumulate} that implements this algorithm together with
some utility functions that make working with collapsing schemes easier.

@@

\section{The accumulate R package}
Grouped aggregation with a fall-though scenario based on a collapsing scheme
requires a lot of parameters to be specified by the user. Besides the data, one
needs to specify the method(s) of aggregation, the collapsing scheme, and the
condition to decide whether a subset is fit for aggregation or a next collapse
is necessary. The main function of the package is caled \texttt{collapse}
and has the following signature.
\begin{verbatim}
  accumulate(data, collapse, test, ...)
\end{verbatim}
where \texttt{data} is a data frame, \texttt{collapse} represents the collapse
sequence, \texttt{test} is a function that accepts a subset of \texttt{data}
and returns \texttt{TRUE} or \texttt{FALSE} and the ellepsis is a sequence
of \texttt{label = expression} pairs that represent aggregation
functions, in a similar style as \texttt{dplyr::summarize}. The output
is a data frame that can schematically be represented as follows.
\begin{verbatim}
  [Grouping Variables, Collapse level, Output aggregates]
\end{verbatim}
The first columns represent the variables that define the output grouping, the
next column is an integer that indicates the level of collapsing used to
compute the aggregate, and the last set of columns store the aggregates.

@@

\section{Extensive example: imputing economic data}

\section{Summary and conclusion}

>>>>

In all packages mentioned, the calculation for each group uses data available
within the group. However, there are valid use cases where a group aggregate is
determined using attributes from out-of-group entities. One example where this
occurs is in the area of small area estimation (see e.g.
\citet{molina2015sae}. Here, one wishes to estimate an aggregate for a group,
for example a geographical region, or a detailed population subset, where the
number of (sampled) observations so small that the variance of the estimate
would be unacceptably large. In small area estimation (SAE) one trades bias for
variance by `borrowing statistical strength' from out-of-group records. The
out-of-group records can be obtained, for example by combining the original
small group with a group of records that are deemed similar in certain
respects. A second area where out-of-group records play a role is in certain
hot-deck imputation methods \citep{andridge2010review}.  In
$k$-nearest-neighbours imputation for example, one finds a set of $k$ donor
records that are preferably in the same group, but this condition may be
relaxed if there are not enough records in the group.  In the \pkg{VIM} package
for \proglang{R} \citep{kowarik2016imputation}, this is controlled by a
combination of the Gower distance and setting conditions on minimal number of
donors. In practice, imputation is often performed via a fall-though scenario,
where one first tries to estimate a model within a group, but if the group is
too small for a reliable estimate of model parameters, the group is enlarged by
combining similar groups, similar to the small-area estimation scenario. A
third example comes from the field of statistical disclosure control (SDC)
\citep{hundepool2012statistical}. In this field, the so-called $p$\% rule
states that the value in a grouped aggregation may not be determined for more
than $p$ percent by a single entity. This occurs for example, when a certain
economic activity in a geographic region is dominated by a single company. In
such cases the value can either not be published (suppression), or one
publishes an aggregate over larger groups. Similarly for microdata there is the
concept of $k$-anonymity, which states that each combination of properties must
occur at least $k$ times in a data set so as to prevent reidentification or
disclosure of individual properties. One way to achieve this is to combine
detailed groupings based on categorical variables into larger groups. For
example by combining groups that in a hierarchical classification have the same
parent.

Both SAE and SDC are supported by \proglang{R} and other free software. Small
area estimation methodology has for example been implemented in the \pkg{sae}
package of \cite{molina2015sae} and the \pkg{hbsae} package of
\cite{boonstra2022hbsae}.  Disclosure control methodology is implemented in the
free and open source standalone software $\tau-$Argus for tabular data and
$\mu$-Argus for microdata \citep{argus2022}, and is also supported by
\proglang{R} packages \pkg{sdcMicro} and \pkg{sdcTable}
\citep{templ2015statistical, meindl2022sdctable}. Regarding imputation
methodology, the CRAN task view on missing
data\footnote{https://cran.r-project.org/web/views/MissingData.html} currently
lists 203 \pkg{R} packages that support some form of estimating missing data.
The \pkg{simputation} package \citep{loo2022simputation} seems to be the only
one that allows for some kind of fall-through scenario for selecting methods,
but it does not allow for dynamic grouping.

Summarizing, we see that on one hand there are many implementations available
for generic aggregation based on fixed groups. On the other hand there are
domain-specific implementation for methods where dynamic grouping of a set of
records plays a role. This paper presents a generic solution to
split-apply-combine aggregation where groups can be collapsed dynamically in
the form of an \pkg{R} package \pkg{accumulate} \cite{loo2022accumulate}.

The \code{accumulate} package serves the use case where a user wishes to
compute aggregates for a certain grouping of records. However, if a certain
instance of a group does not meet user-defined quality requirements, the set of
records is expanded by (recursively) collapsing the grouping according to a
user-defined scheme. For example, given some financial data on companies, one
wishes to compute the average profit to turnover ratio for each combination of
economic activity and size class. If for a certain combination of economic
activity and size class there are too few records for a reliable estimate, one
could drop size class and compute the average ratio over all records within a
certain economic activity. Or, one could choose to course-grain economic
activities by collapsing groups with activities that are deemed similar enough.

The package has been developed with the following design choices in mind.
First, the interface should be easy to learn for \proglang{R} users, and thus
should resemble existing popular interfaces where possible. Second, users
should be free to define any possibly multi-step, collapsing scheme.  Here, we
keep in mind that collapsing schemes may be constructed manually based on
domain knowledge and that users may want to experiment with several schemes
before deciding on a final solution. This calls for a certain separation of
concerns between defining collapsing schemes and applying them to data.  The
package should also support collapsing schemes that follow naturally from
hierarchical classification systems. Third, users should have the flexibility
to define any quality requirement on the grouping while common quality
requiremtns are supported out of the box. Common quality requirements include a
minimum number of records, or a minum fraction of records without missing
values, or a minimum number of records with non-zero values.

The rest of this paper is organized as follows. In the next section we formally
define the concepts of the quality requirements and a collapsing scheme. We
describe the fundamental algorithm in pseudocode and compare it with the
standard grouping scenario. We also discuss the time and space complexity and
some caveats with implementation and performance. This section is mainly of
interest for developers who wish to include such functionality in their own
package. In Section~\ref{sect:accumulate} we describe \pkg{accumulate} package
and demonstrate how the design choices have determined its implementation.  We
give some toy examples to extensively discuss the input and output of the main
functions, discuss how to define tests in some detail, and also explain how one
can program over the functions in the package.  Section~\ref{sect:example}
works through a realistic example based on a synthetic dataset of economic
microdata. We conclude with a summary and outlook in
Section~\ref{sect:conclusion}.

@@

\section{The accumulate R package}\label{sect:accumulate}
Grouped aggregation with a fall-though scenario based on a collapsing scheme
requires a lot of parameters to be specified by the user. Besides the data, one
needs to specify the method(s) of aggregation, the collapsing scheme, and the
condition to decide whether a subset is fit for aggregation or a next collapse
is necessary. The main function of the package is caled \texttt{collapse}
and has the following signature.
\begin{verbatim}
  accumulate(data, collapse, test, ...)
\end{verbatim}
where \texttt{data} is a data frame, \texttt{collapse} represents the collapse
sequence, \texttt{test} is a function that accepts a subset of \texttt{data}
and returns \texttt{TRUE} or \texttt{FALSE} and the ellepsis is a sequence
of \texttt{label = expression} pairs that represent aggregation
functions, in a similar style as \texttt{dplyr::summarize}. The output
is a data frame that can schematically be represented as follows.
\begin{verbatim}
  [Grouping Variables, Collapse level, Output aggregates]
\end{verbatim}
The first columns represent the variables that define the output grouping, the
next column is an integer that indicates the level of collapsing used to
compute the aggregate, and the last set of columns store the aggregates.

@@

\section{Extensive example: imputing economic data} \label{sect:example}

\section{Summary and conclusion}\label{sect:conclusion}
